---
title: "Data Visualization"
format:
  html:
    toc: true
    toc-depth: 3
    link-external-newwindow: true
    page-layout: full
code-fold: true
code-summary: "Show Code"
execute:
  warning: false
  message: false
editor : source
---

## Dataset Overview


This comprehensive analysis examines U.S. inflation trends through the lens of everyday American experiences, using Consumer Price Index (CPI) data from the Federal Reserve Economic Data (FRED) spanning January 2010 to December 2024. Our investigation begins with food inflation—the most immediate and visceral form of price pressure that families face daily. The original dataset comprises four key food price indices: Food at Home CPI (representing grocery costs), Food Away from Home CPI (restaurant and dining costs), Food Total CPI (aggregate food prices), and Overall CPI (economy-wide price levels). These monthly time series capture approximately 15 years of dining table economics, encompassing multiple economic cycles including the 2008 financial crisis recovery, pre-pandemic stability, COVID-19 disruption, and subsequent policy responses.

Building upon this foundation, we've expanded our analysis to encompass the broader landscape of American cost of living through three additional critical dimensions that shape household budgets and financial well-being. Housing costs represent the largest expense for most families, tracked through Shelter CPI (overall housing expenses), Rent CPI (rental market dynamics), and 30-Year Mortgage Rates (the gateway to homeownership). Energy prices, captured through Gasoline CPI, serve as both a direct household expense and a barometer of broader economic volatility that ripples through transportation, goods delivery, and consumer sentiment. Finally, the financial environment—measured by Federal Funds Rate and 10-Year Treasury Rate—reveals how monetary policy decisions in Washington boardrooms ultimately affect family budgets across America.

Together, these indicators weave a story of how abstract economic forces translate into concrete realities: the monthly rent check, the weekly grocery bill, the daily commute, and the long-term dream of homeownership. This multi-dimensional approach captures not just what inflation looks like in government statistics, but how it feels in American households.



```{r}

# Load libraries
library(tidyverse)
library(lubridate)
library(plotly)
library(ggplot2)
library(forecast)
library(tseries)
library(gridExtra)
library(viridis)
library(httr)
library(jsonlite)
library(dplyr)
library(readr)
library(scales)
library(tidyr)

# Read all CSV files
shelter_cpi <- read_csv("./data/CUSR0000SAH1.csv")
rent_cpi <- read_csv("./data/CUSR0000SEHA.csv")
mortgage_rate <- read_csv("./data/MORTGAGE30US.csv")
gasoline_cpi <- read_csv("./data/CUSR0000SETB01.csv")
treasury_10y <- read_csv("./data/DGS10.csv")
fed_funds <- read_csv("./data/FEDFUNDS.csv")

shelter_cpi$DATE <- as.Date(shelter_cpi$observation_date)
rent_cpi$DATE <- as.Date(rent_cpi$observation_date)
mortgage_rate$DATE <- as.Date(mortgage_rate$observation_date)
gasoline_cpi$DATE <- as.Date(gasoline_cpi$observation_date)
treasury_10y$DATE <- as.Date(treasury_10y$observation_date)
fed_funds$DATE <- as.Date(fed_funds$observation_date)

# Rename value columns for clarity (assuming second column contains values)
names(shelter_cpi)[2] <- "Shelter_CPI"
names(rent_cpi)[2] <- "Rent_CPI"
names(mortgage_rate)[2] <- "Mortgage_Rate"
names(gasoline_cpi)[2] <- "Gasoline_CPI"
names(treasury_10y)[2] <- "Treasury_10Y"
names(fed_funds)[2] <- "Fed_Funds_Rate"

# Set theme
theme_set(theme_minimal() + 
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10)
  ))


# Create data directory
if (!dir.exists("data")) {
  dir.create("data")
  cat("Created 'data' directory\n")
}

# Food-related FRED series IDs 
food_series <- list(
  list(id = "CUSR0000SAF11", name = "Food at Home CPI"),
  list(id = "CUSR0000SEFV", name = "Food Away from Home CPI"),
  list(id = "CUSR0000SAF1", name = "Food and Beverages Total CPI"),
  list(id = "CPIAUCSL", name = "Overall CPI (All Items)")
)

# Download all food-related data
food_data_list <- list()

# Check what data we successfully downloaded
successful_downloads <- names(food_data_list)
for (id in successful_downloads) {
  series_name <- food_series[[which(sapply(food_series, function(x) x$id == id))]]$name
  cat("✓", id, "-", series_name, "\n")
}

# Function to load and clean FRED CSV files
load_fred_csv <- function(file_path, series_name) {

  tryCatch({
    # Read the CSV file
    data <- read_csv(file_path, show_col_types = FALSE)
    
    # Handle different CSV formats from FRED
    if (ncol(data) >= 2) {
      # Check column names and standardize
      col_names <- names(data)
      
      # Find date column
      date_col <- col_names[grepl("DATE|date|Date", col_names)][1]
      if (is.na(date_col)) date_col <- col_names[1]  # First column if no DATE found
      
      # Find value column  
      value_col <- col_names[!col_names %in% date_col][1]  # First non-date column
      
      # Select and rename columns
      data <- data %>% 
        select(date = all_of(date_col), value = all_of(value_col))
      
      # Clean and convert data
      data <- data %>%
        mutate(
          date = as.Date(date),
          value = as.numeric(as.character(value))
        ) %>%
        filter(!is.na(date), !is.na(value), value != ".", !is.infinite(value)) %>%
        arrange(date)
      
      return(data)
    } else {
      cat("✗ Unexpected file format for", series_name, "\n")
      return(NULL)
    }
    
  }, error = function(e) {
    cat("✗ Error loading", series_name, ":", e$message, "\n")
    return(NULL)
  })
}

# Define the data files we expect
food_data_files <- list(
  list(path = "data/CUSR0000SAF11.csv", name = "Food at Home CPI", var_name = "food_at_home"),
  list(path = "data/CUSR0000SEFV.csv", name = "Food Away from Home CPI", var_name = "food_away_home"),
  list(path = "data/CPIAUCSL.csv", name = "Overall CPI (All Items)", var_name = "cpi_all")
)

# Check for additional files
if (file.exists("data/CUSR0000SAF1.csv")) {
  food_data_files <- append(food_data_files, 
    list(list(path = "data/CUSR0000SAF1.csv", name = "Food and Beverages Total CPI", var_name = "food_total")), 
    after = 2)
}

# Load all available files
loaded_data <- list()
for (i in 1:length(food_data_files)) {
  file_info <- food_data_files[[i]]
  
  if (file.exists(file_info$path)) {
    data <- load_fred_csv(file_info$path, file_info$name)
    if (!is.null(data)) {
      # Rename the value column to the appropriate variable name
      data <- data %>% rename(!!file_info$var_name := value)
      loaded_data[[file_info$var_name]] <- data
    }
  } else {
    cat("File not found:", file_info$path, "\n")
  }
}

# Combine all loaded data
if (length(loaded_data) >= 2) {
  # Start with the first dataset
  combined_food_data <- loaded_data[[1]]
  
  # Join other datasets
  for (i in 2:length(loaded_data)) {
    combined_food_data <- combined_food_data %>%
      full_join(loaded_data[[i]], by = "date")
  }
  
  # If we don't have food_total but have both components, calculate it
  if (!"food_total" %in% names(combined_food_data) && 
      all(c("food_at_home", "food_away_home") %in% names(combined_food_data))) {
    
    combined_food_data <- combined_food_data %>%
      mutate(food_total = (food_at_home + food_away_home) / 2)
  }
  
  # Sort by date and remove rows with all missing values
  combined_food_data <- combined_food_data %>%
    arrange(date) %>%
    filter(if_any(-date, ~ !is.na(.x)))

  
} else {
  stop("ERROR: Need at least 2 data series to proceed. Please check your data files in the 'data/' folder.")
}

# Summary statistics
data_summary <- combined_food_data %>%
  summarise(
    observations = n(),
    date_start = min(date, na.rm = TRUE),
    date_end = max(date, na.rm = TRUE),
    years_covered = round(as.numeric(difftime(max(date, na.rm = TRUE), min(date, na.rm = TRUE), units = "days")) / 365.25, 1)
  )

# Check completeness for each variable
available_vars <- names(combined_food_data)[names(combined_food_data) != "date"]
completeness <- combined_food_data %>%
  summarise(across(all_of(available_vars), ~ round(sum(!is.na(.x)) / n() * 100, 1), .names = "{.col}_complete"))

```

## Food

### Interactive Food Inflation Dashboard

The comprehensive dashboard reveals distinct phases in food price evolution through multiple analytical lenses. The upper left panel displays absolute CPI levels, showing steady upward trajectories with notable acceleration post-2020. Food away from home (orange line) exhibits the steepest price growth, reaching nearly 380 index points by 2024, while food at home (blue line) demonstrates more moderate increases. 

The year-over-year inflation rates (upper right) expose the volatile nature of food price changes, with dramatic spikes during 2021-2022 reaching 10-14% annually—levels not observed since the early 1980s. The indexed comparison (lower left) normalizes all series to 100 at the start period, clearly illustrating that restaurant prices have increased approximately 65% compared to 45% for groceries since 2010. The recent three-year trend (lower right) captures the inflation surge and subsequent moderation, with most categories returning toward 2-4% annual growth by late 2024.


```{r}
#| label: plotly-dashboard
#| fig-cap: "Food Inflation Dashboard"
#| out-width: "100%"
create_plotly_dashboard <- function(data) {
  
  # Get available columns (excluding date)
  available_cols <- names(data)[names(data) != "date"]
  
  # Calculate year-over-year inflation rates
  inflation_data <- data %>%
    arrange(date) %>%
    mutate(
      across(all_of(available_cols), 
             ~ (. / lag(., 12) - 1) * 100, 
             .names = "{.col}_yoy")
    ) %>%
    filter(if_any(ends_with("_yoy"), ~ !is.na(.x)))
  
  # Color palette
  colors <- c('#1f77b4', '#ff7f0e', '#2ca02c', '#d62728', '#9467bd')
  
  # PLOT 1: Price Index Levels
  p1 <- plot_ly(type = 'scatter', mode = 'lines')
  
  for (i in seq_along(available_cols)) {
    col_name <- available_cols[i]
    display_name <- case_when(
      col_name == "food_at_home" ~ "Food at Home",
      col_name == "food_away_home" ~ "Food Away from Home",
      col_name == "food_total" ~ "Food Total",
      col_name == "cpi_all" ~ "Overall CPI",
      TRUE ~ str_to_title(str_replace_all(col_name, "_", " "))
    )
    
    p1 <- p1 %>% 
      add_trace(data = data, x = ~date, y = data[[col_name]], 
                name = display_name,
                line = list(color = colors[i], width = 2),
                hovertemplate = paste('<b>', display_name, '</b><br>',
                                     'Date: %{x}<br>',
                                     'Value: %{y:.1f}<br>',
                                     '<extra></extra>'))
  }
  
  p1 <- p1 %>%
    layout(
      title = "Price Index Levels",
      xaxis = list(title = "Date"),
      yaxis = list(title = "Index Value"),
      showlegend = TRUE,
      shapes = list(
        list(
          type = "line",
          x0 = as.Date("2020-03-01"), x1 = as.Date("2020-03-01"),
          y0 = 0, y1 = 1,
          yref = "paper",
          line = list(color = "red", dash = "dot", width = 1)
        )
      )
    )
  
  # PLOT 2: YoY Inflation Rates
  p2 <- plot_ly(type = 'scatter', mode = 'lines')
  
  yoy_cols <- paste0(available_cols, "_yoy")
  for (i in seq_along(yoy_cols)) {
    col_name <- yoy_cols[i]
    if (col_name %in% names(inflation_data)) {
      orig_name <- available_cols[i]
      display_name <- case_when(
        orig_name == "food_at_home" ~ "Food at Home",
        orig_name == "food_away_home" ~ "Food Away",
        orig_name == "food_total" ~ "Food Total",
        orig_name == "cpi_all" ~ "Overall CPI",
        TRUE ~ str_to_title(str_replace_all(orig_name, "_", " "))
      )
      
      p2 <- p2 %>% 
        add_trace(data = inflation_data, x = ~date, y = inflation_data[[col_name]], 
                  name = paste(display_name, "YoY%"),
                  line = list(color = colors[i], width = 2),
                  hovertemplate = paste('<b>', display_name, '</b><br>',
                                       'Date: %{x}<br>',
                                       'YoY Change: %{y:.1f}%<br>',
                                       '<extra></extra>'))
    }
  }
  
  p2 <- p2 %>%
    layout(
      title = "Year-over-Year Inflation Rates (%)",
      xaxis = list(title = "Date"),
      yaxis = list(title = "YoY % Change"),
      showlegend = TRUE,
      shapes = list(
        list(
          type = "line",
          x0 = 0, x1 = 1,
          xref = "paper",
          y0 = 0, y1 = 0,
          line = list(color = "gray", dash = "dash", width = 1)
        ),
        list(
          type = "line",
          x0 = as.Date("2020-03-01"), x1 = as.Date("2020-03-01"),
          y0 = 0, y1 = 1,
          yref = "paper",
          line = list(color = "red", dash = "dot", width = 1)
        )
      )
    )
  
  # PLOT 3: Indexed trends (first value = 100)
  indexed_data <- data %>%
    mutate(across(all_of(available_cols), 
                  ~ (. / first(., na_rm = TRUE)) * 100, 
                  .names = "{.col}_idx"))
  
  p3 <- plot_ly(type = 'scatter', mode = 'lines')
  
  idx_cols <- paste0(available_cols, "_idx")
  for (i in seq_along(idx_cols)) {
    col_name <- idx_cols[i]
    orig_name <- available_cols[i]
    display_name <- case_when(
      orig_name == "food_at_home" ~ "Food at Home",
      orig_name == "food_away_home" ~ "Food Away",
      orig_name == "food_total" ~ "Food Total",
      orig_name == "cpi_all" ~ "Overall CPI",
      TRUE ~ str_to_title(str_replace_all(orig_name, "_", " "))
    )
    
    p3 <- p3 %>% 
      add_trace(data = indexed_data, x = ~date, y = indexed_data[[col_name]], 
                name = paste(display_name, "(Index)"),
                line = list(color = colors[i], width = 2),
                hovertemplate = paste('<b>', display_name, '</b><br>',
                                     'Date: %{x}<br>',
                                     'Index: %{y:.1f}<br>',
                                     '<extra></extra>'))
  }
  
  p3 <- p3 %>%
    layout(
      title = "Cumulative Price Changes (First Value = 100)",
      xaxis = list(title = "Date"),
      yaxis = list(title = "Index (Base = 100)"),
      showlegend = TRUE,
      shapes = list(
        list(
          type = "line",
          x0 = 0, x1 = 1,
          xref = "paper",
          y0 = 100, y1 = 100,
          line = list(color = "gray", dash = "dash", width = 1)
        ),
        list(
          type = "line",
          x0 = as.Date("2020-03-01"), x1 = as.Date("2020-03-01"),
          y0 = 0, y1 = 1,
          yref = "paper",
          line = list(color = "red", dash = "dot", width = 1)
        )
      )
    )
  
  # PLOT 4: Recent trends (last 3 years)
  recent_data <- inflation_data %>%
    filter(date >= max(date, na.rm = TRUE) - years(3))
  
  p4 <- plot_ly(type = 'scatter', mode = 'lines+markers')
  
  if (nrow(recent_data) > 0) {
    for (i in seq_along(yoy_cols)) {
      col_name <- yoy_cols[i]
      if (col_name %in% names(recent_data)) {
        orig_name <- available_cols[i]
        display_name <- case_when(
          orig_name == "food_at_home" ~ "Food at Home",
          orig_name == "food_away_home" ~ "Food Away",
          orig_name == "food_total" ~ "Food Total", 
          orig_name == "cpi_all" ~ "Overall CPI",
          TRUE ~ str_to_title(str_replace_all(orig_name, "_", " "))
        )
        
        p4 <- p4 %>% 
          add_trace(data = recent_data, x = ~date, y = recent_data[[col_name]], 
                    name = display_name,
                    line = list(color = colors[i], width = 2),
                    marker = list(size = 4, color = colors[i]),
                    hovertemplate = paste('<b>', display_name, '</b><br>',
                                         'Date: %{x}<br>',
                                         'YoY Change: %{y:.1f}%<br>',
                                         '<extra></extra>'))
      }
    }
  }
  
  p4 <- p4 %>%
    layout(
      title = "Recent 3-Year Inflation Trend",
      xaxis = list(title = "Date"),
      yaxis = list(title = "YoY % Change"),
      showlegend = TRUE,
      shapes = list(
        list(
          type = "line",
          x0 = 0, x1 = 1,
          xref = "paper",
          y0 = 0, y1 = 0,
          line = list(color = "gray", dash = "dash", width = 1)
        )
      )
    )
  combined_plot <- subplot(
    subplot(p1, p2, nrows = 1, shareY = FALSE, titleX = TRUE, titleY = TRUE),
    subplot(p3, p4, nrows = 1, shareY = FALSE, titleX = TRUE, titleY = TRUE),
    nrows = 2, shareX = FALSE, titleX = TRUE, titleY = TRUE,
    heights = c(0.5, 0.5)
  ) %>%
  layout(
    title = list(
      text = "<b>Food Inflation Dashboard: Real CPI Data Analysis</b><br><sup>Interactive analysis of food price trends and inflation rates | Red dashed lines mark COVID-19 onset (March 2020)</sup>",
      font = list(size = 16)
    ),
    template = "plotly_white",
    height = 800,
    showlegend = FALSE
  )
  
  return(combined_plot)
}

plotly_dashboard <- create_plotly_dashboard(combined_food_data)
plotly_dashboard
```

### Seasonal and Period-Based Inflation Patterns

The heatmap analysis reveals systematic differences across time periods and seasonal patterns. The COVID era (2020-2022) displays predominantly red coloration, indicating sustained high inflation across all food categories throughout most months. This contrasts sharply with the pre-COVID period (2011-2019), characterized by predominantly neutral to mild positive inflation rates. The post-COVID era (2023+) shows a cooling pattern, with inflation rates moderating but remaining elevated compared to historical norms.

Seasonal patterns emerge within each period: food prices typically show modest increases during winter months and more variable behavior during summer periods. The intensity of seasonal effects appears amplified during the COVID era, suggesting that supply chain disruptions interacted with normal seasonal variations to create more pronounced price swings.



```{r}
# GGPLOT2 Heatmap - Food Inflation Analysis
create_ggplot_heatmap <- function(data) {
  
  # Get available columns
  available_cols <- names(data)[names(data) != "date"]
  
  # Calculate year-over-year changes
  heatmap_data <- data %>%
    arrange(date) %>%
    mutate(
      across(all_of(available_cols), 
             ~ (. / lag(., 12) - 1) * 100, 
             .names = "{.col}_yoy")
    ) %>%
    filter(if_any(ends_with("_yoy"), ~ !is.na(.x))) %>%
    select(date, ends_with("_yoy")) %>%
    mutate(
      year = year(date),
      month = month(date, label = TRUE, abbr = TRUE)
    ) %>%
    pivot_longer(cols = ends_with("_yoy"), names_to = "category", values_to = "inflation_rate") %>%
    mutate(
      category_clean = case_when(
        category == "food_at_home_yoy" ~ "Food at Home",
        category == "food_away_home_yoy" ~ "Food Away from Home",
        category == "food_total_yoy" ~ "Food Total",
        category == "cpi_all_yoy" ~ "Overall CPI",
        TRUE ~ str_remove(category, "_yoy") %>% str_to_title()
      ),
      period = case_when(
        year <= 2019 ~ "Pre-COVID (2011-2019)",
        year %in% 2020:2022 ~ "COVID Era (2020-2022)",
        year >= 2023 ~ "Post-COVID (2023+)",
        TRUE ~ "Other"
      )
    ) %>%
    filter(!is.na(inflation_rate))
  
  if (nrow(heatmap_data) == 0) {
    return(ggplot() + theme_void() + labs(title = "Insufficient data for heatmap"))
  }
  
  # Create the heatmap with proper dimensions and clean layout
  g <- ggplot(heatmap_data, aes(x = month, y = fct_rev(category_clean), fill = inflation_rate)) +
    geom_tile(color = "white", size = 0.8) +
    # Remove ALL text inside tiles for clean appearance
    scale_fill_gradient2(
      low = "#2166AC", mid = "white", high = "#B2182B",
      midpoint = 0, name = "YoY %\nChange",
      limits = c(-10, 20), oob = scales::squish,
      breaks = seq(-10, 20, by = 5)
    ) +
    facet_wrap(~ period, ncol = 1, strip.position = "top") +  # Single column layout
    labs(
      title = "Food Inflation Heatmap: Year-over-Year Changes",
      subtitle = "Seasonal patterns across different time periods",
      x = "Month",
      y = "Food Category",
      caption = "Data: Federal Reserve Economic Data (FRED)"
    ) +
    theme_minimal() +
    theme(
      # Fix text sizing and spacing
      axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5),
      axis.text.y = element_text(size = 10),
      
      # Clean facet labels
      strip.background = element_rect(fill = "lightgray", alpha = 0.8),
      strip.text = element_text(face = "bold", size = 11),
      
      # Title formatting
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5),
      plot.caption = element_text(size = 10, hjust = 0.5),
      
      # Legend positioning
      legend.position = "right",
      legend.title = element_text(size = 11),
      legend.text = element_text(size = 10),
      
      # Proper spacing
      panel.spacing.y = unit(1.5, "lines"),
      plot.margin = margin(20, 20, 20, 20),
      
      # Remove grid
      panel.grid = element_blank()
    )
  
  return(g)
}

create_ggplot_heatmap_horizontal <- function(data) {
  
  # Get available columns
  available_cols <- names(data)[names(data) != "date"]
  
  # Calculate year-over-year changes 
  heatmap_data <- data %>%
    arrange(date) %>%
    mutate(
      across(all_of(available_cols), 
             ~ (. / lag(., 12) - 1) * 100, 
             .names = "{.col}_yoy")
    ) %>%
    filter(if_any(ends_with("_yoy"), ~ !is.na(.x))) %>%
    select(date, ends_with("_yoy")) %>%
    mutate(
      year = year(date),
      month = month(date, label = TRUE, abbr = TRUE)
    ) %>%
    pivot_longer(cols = ends_with("_yoy"), names_to = "category", values_to = "inflation_rate") %>%
    mutate(
      category_clean = case_when(
        category == "food_at_home_yoy" ~ "Food at Home",
        category == "food_away_home_yoy" ~ "Food Away from Home",
        category == "food_total_yoy" ~ "Food Total",
        category == "cpi_all_yoy" ~ "Overall CPI",
        TRUE ~ str_remove(category, "_yoy") %>% str_to_title()
      ),
      period = case_when(
        year <= 2019 ~ "Pre-COVID\n(2011-2019)",
        year %in% 2020:2022 ~ "COVID Era\n(2020-2022)",
        year >= 2023 ~ "Post-COVID\n(2023+)",
        TRUE ~ "Other"
      )
    ) %>%
    filter(!is.na(inflation_rate))
  
  if (nrow(heatmap_data) == 0) {
    return(ggplot() + theme_void() + labs(title = "Insufficient data for heatmap"))
  }
  
  # Horizontal layout version
  g <- ggplot(heatmap_data, aes(x = month, y = fct_rev(category_clean), fill = inflation_rate)) +
    geom_tile(color = "white", size = 0.5) +
    scale_fill_gradient2(
      low = "#2166AC", mid = "white", high = "#B2182B",
      midpoint = 0, name = "YoY %\nChange",
      limits = c(-10, 20), oob = scales::squish
    ) +
    facet_wrap(~ period, ncol = 3) +  
    labs(
      title = "Food Inflation Heatmap by Period",
      x = "Month",
      y = "Food Category",
      caption = "Source: Federal Reserve Economic Data (FRED)"
    ) +
    theme_minimal() +
    theme(
      # Larger text for better readability
      axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
      axis.text.y = element_text(size = 10),
      
      # Better facet formatting
      strip.background = element_rect(fill = "lightgray", alpha = 0.7),
      strip.text = element_text(face = "bold", size = 10),
      
      # Clean titles
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      axis.title = element_text(size = 11),
      
      # Legend on bottom to save horizontal space
      legend.position = "bottom",
      legend.title = element_text(size = 10),
      
      # Better spacing
      panel.spacing = unit(1, "lines"),
      plot.margin = margin(15, 15, 15, 15),
      panel.grid = element_blank()
    ) +
    guides(fill = guide_colorbar(title.position = "top", title.hjust = 0.5, barwidth = 15, barheight = 1))
  
  return(g)
}

# Use the horizontal version for better display
ggplot_heatmap <- create_ggplot_heatmap_horizontal(combined_food_data)
print(ggplot_heatmap)

```
### Price Volatility Analysis

The volatility analysis quantifies price uncertainty through rolling 12-month standard deviations of monthly price changes. Food at home (blue line) exhibits the highest volatility throughout most of the analysis period, with dramatic spikes reaching 0.8% monthly volatility during 2020-2021. This represents unprecedented price unpredictability, far exceeding levels observed during the 2008-2009 financial crisis.

The COVID-19 period marks a clear structural break in price stability across all food categories. Pre-pandemic volatility typically ranged between 0.1-0.3% monthly, while post-COVID levels reached 0.4-0.8%. The synchronized increase across all categories suggests systemic rather than category-specific shocks. By 2024, volatility has declined but remains above pre-pandemic baselines, indicating persistent supply chain fragility.

```{r}
create_ggplot_volatility <- function(data) {
  
  # Get available columns
  available_cols <- names(data)[names(data) != "date"]
  
  if (length(available_cols) == 0) {
    return(ggplot() + theme_void() + labs(title = "No data available for volatility analysis"))
  }
  
  # Calculate price changes and rolling volatility
  volatility_data <- data %>%
    arrange(date) %>%
    mutate(
      # Monthly price changes
      across(all_of(available_cols), 
             ~ (. / lag(.) - 1) * 100, 
             .names = "{.col}_change")
    )
  
  # Calculate rolling volatility manually for each change column
  change_cols <- paste0(available_cols, "_change")
  
  for (col in change_cols) {
    if (col %in% names(volatility_data)) {
      # Use rollapply instead of rollsd
      volatility_data[[paste0(col, "_vol")]] <- zoo::rollapply(
        volatility_data[[col]], 
        width = 12, 
        FUN = function(x) sd(x, na.rm = TRUE), 
        fill = NA, 
        align = "right",
        partial = FALSE
      )
    }
  }
  
  # Continue with the rest of the analysis
  volatility_data <- volatility_data %>%
    select(date, ends_with("_vol")) %>%
    filter(if_any(ends_with("_vol"), ~ !is.na(.x))) %>%
    pivot_longer(cols = -date, names_to = "category", values_to = "volatility") %>%
    mutate(
      category_clean = case_when(
        str_detect(category, "food_at_home") ~ "Food at Home",
        str_detect(category, "food_away_home") ~ "Food Away from Home",
        str_detect(category, "food_total") ~ "Food Total", 
        str_detect(category, "cpi_all") ~ "Overall CPI",
        TRUE ~ str_remove_all(category, "_change_vol") %>% str_to_title()
      )
    ) %>%
    filter(!is.na(volatility))
  
  if (nrow(volatility_data) == 0) {
    return(ggplot() + theme_void() + labs(title = "Insufficient data for volatility analysis"))
  }
  
  # Select colors based on available categories
  n_categories <- length(unique(volatility_data$category_clean))
  colors <- c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd")[1:n_categories]
  
  # Create the volatility plot
  g <- ggplot(volatility_data, aes(x = date, y = volatility, color = category_clean)) +
    geom_line(size = 1.2, alpha = 0.8) +
    geom_smooth(method = "loess", span = 0.3, se = TRUE, alpha = 0.2, size = 0.8) +
    scale_color_manual(values = colors, name = "Category") +
    labs(
      title = "Food Price Volatility: Rolling 12-Month Standard Deviation",
      subtitle = "Measuring price stability across food categories over time",
      x = "Date",
      y = "Monthly Price Change Volatility (%)",
      caption = "Higher volatility = more unpredictable price changes | Smooth lines show trends"
    ) +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      plot.title = element_text(size = 16, face = "bold"),
      plot.subtitle = element_text(size = 12)
    ) +
    geom_vline(xintercept = as.Date("2020-03-01"), linetype = "dashed", color = "red", alpha = 0.7) +
    geom_vline(xintercept = as.Date("2008-09-01"), linetype = "dashed", color = "blue", alpha = 0.7) +
    annotate("text", x = as.Date("2020-06-01"), y = max(volatility_data$volatility, na.rm = TRUE) * 0.9, 
             label = "COVID-19", color = "red", size = 3.5) 
  
  return(g)
}

# Create and display volatility analysis
ggplot_volatility <- create_ggplot_volatility(combined_food_data)
print(ggplot_volatility)
```

In terms of the food prices part, the data narrative reveals a three-act inflation story. Act I (2010-2019) represents the "Great Moderation" period with stable, predictable food price growth averaging 1-2% annually. Act II (2020-2022) depicts the "Great Disruption"—unprecedented supply chain shocks, monetary expansion, and demand shifts creating multi-decade high inflation rates. Act III (2023-2024) shows "Gradual Normalization"—inflation cooling but price levels permanently elevated, with volatility settling at new, higher equilibrium levels.

The March 2020 inflection point visible across all visualizations marks when pandemic-induced disruptions transformed food markets. Supply chain bottlenecks, labor shortages, and shifting consumption patterns from restaurants to groceries created sustained inflationary pressures. The differential impact between food at home and away from home reflects these consumption shifts and varying supply chain vulnerabilities.

This demonstrates how external shocks can fundamentally alter inflation dynamics, with effects persisting long after initial disruptions subside. The food sector's essential nature amplified both the immediate impacts and policy responses, creating feedback loops that sustained elevated price growth well into the recovery period.


## Housing: When Home Becomes Unaffordable

The housing market tells perhaps the most dramatic story of the past decade, one that transforms the American Dream into an increasingly distant aspiration for millions. What we see in the data is not just a gradual price increase, but a fundamental restructuring of housing affordability that began with the pandemic and shows little sign of reversing. The steady climb of shelter costs from 2010 to 2020 represented a manageable, if concerning, trend. Families could plan, save, and reasonably expect to eventually afford homeownership or rental upgrades.

Then came March 2020, and everything changed. The pandemic didn't just accelerate existing trends—it shattered them entirely. Housing costs began climbing at rates that outpaced income growth by unprecedented margins. The Federal Reserve's decision to slash interest rates to near zero, while necessary to prevent economic collapse, inadvertently poured gasoline on the housing fire. Cheap money flowed into real estate markets just as changing work patterns drove families to seek larger homes in different locations. The result was a perfect storm that priced out an entire generation of potential homebuyers and renters alike.

```{r}

# Filter data from 2000 onwards for better storytelling
start_date <- as.Date("2000-01-01")
shelter_cpi <- shelter_cpi[shelter_cpi$DATE >= start_date, ]
rent_cpi <- rent_cpi[rent_cpi$DATE >= start_date, ]
mortgage_rate <- mortgage_rate[mortgage_rate$DATE >= start_date, ]
gasoline_cpi <- gasoline_cpi[gasoline_cpi$DATE >= start_date, ]
treasury_10y <- treasury_10y[treasury_10y$DATE >= start_date, ]
fed_funds <- fed_funds[fed_funds$DATE >= start_date, ]

# Define key economic events for annotations
economic_events <- data.frame(
  date = as.Date(c("2001-09-11", "2008-09-15", "2020-03-15", "2022-03-01")),
  event = c("9/11 Attacks", "Lehman Brothers\nCollapse", "COVID-19\nPandemic", "Russia-Ukraine\nConflict"),
  color = c("#FF6B6B", "#4ECDC4", "#45B7D1", "#96CEB4")
)

data_start <- max(min(shelter_cpi$DATE, na.rm = TRUE), 
                  min(rent_cpi$DATE, na.rm = TRUE), 
                  min(mortgage_rate$DATE, na.rm = TRUE))
data_end <- min(max(shelter_cpi$DATE, na.rm = TRUE), 
                max(rent_cpi$DATE, na.rm = TRUE), 
                max(mortgage_rate$DATE, na.rm = TRUE))

# Combine housing data within actual date range
housing_data <- merge(shelter_cpi, rent_cpi, by = "DATE", all = TRUE) %>%
  merge(mortgage_rate, by = "DATE", all = TRUE) %>%
  filter(DATE >= data_start & DATE <= data_end) %>%
  filter(!is.na(Shelter_CPI) | !is.na(Rent_CPI) | !is.na(Mortgage_Rate))

# Create improved visualization
housing_plot <- ggplot(housing_data, aes(x = DATE)) +
  
  # Housing CPI lines
  geom_line(aes(y = Shelter_CPI, color = "Shelter CPI"), 
            size = 1.5, alpha = 0.9) +
  geom_line(aes(y = Rent_CPI, color = "Rent CPI"), 
            size = 1.5, alpha = 0.9) +
  
  # Mortgage rate on secondary axis 
  geom_line(aes(y = Mortgage_Rate * 12, color = "30-Year Mortgage Rate"), 
            size = 2, alpha = 0.95) +
  
  # COVID-19 period highlight only
  annotate("rect", 
           xmin = as.Date("2020-03-01"), 
           xmax = as.Date("2021-12-31"),
           ymin = -Inf, ymax = Inf, 
           alpha = 0.15, fill = "#3498DB") +
  
  # COVID-19 marker line
  geom_vline(xintercept = as.Date("2020-03-15"), 
             linetype = "dashed", color = "#2C3E50", size = 1, alpha = 0.8) +
  
  # COVID-19 annotation
  annotate("text", 
           x = as.Date("2020-10-01"), 
           y = max(housing_data$Shelter_CPI, na.rm = TRUE) * 0.95, 
           label = "COVID-19\nPandemic Impact", 
           fontface = "bold", size = 2.5, 
           color = "#2C3E50", hjust = 0.5) +
  
  # Dual y-axis setup
  scale_y_continuous(
    name = "Consumer Price Index (CPI)",
    labels = comma_format(),
    sec.axis = sec_axis(~./12, 
                        name = "Mortgage Rate (%)",
                        labels = function(x) paste0(x, "%"))
  ) +
  
  # Color scheme
  scale_color_manual(
    values = c("Shelter CPI" = "#E74C3C", 
               "Rent CPI" = "#3498DB", 
               "30-Year Mortgage Rate" = "#F39C12"),
    name = "Housing Indicators"
  ) +
  
  # X-axis formatting
  scale_x_date(
    date_breaks = "2 years", 
    date_labels = "%Y",
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  
  # Theme and styling
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5, 
                              margin = margin(b = 10)),
    plot.subtitle = element_text(size = 14, hjust = 0.5, 
                                 color = "#34495E", margin = margin(b = 20)),
    legend.position = "bottom",
    legend.title = element_text(face = "bold", size = 12),
    legend.text = element_text(size = 11),
    legend.box.margin = margin(t = 15),
    axis.title = element_text(face = "bold", size = 12),
    axis.title.y.right = element_text(color = "#F39C12"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.text.y = element_text(size = 11),
    axis.text.y.right = element_text(color = "#F39C12", size = 11),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(alpha = 0.3),
    plot.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(20, 20, 20, 20),
    plot.caption = element_text(size = 10, color = "#7F8C8D", hjust = 1)
  ) +
  
  # Labels
  labs(
    title = "Housing Costs Through the Pandemic Era",
    subtitle = "Time-Series Evidence of COVID-19's Impact on U.S. Shelter Affordability",
    x = "Year",
    caption = "Data: Bureau of Labor Statistics & Federal Reserve Economic Data"
  )

# Display the plot with larger optimal size
options(repr.plot.width = 16, repr.plot.height = 10)
print(housing_plot)

```


## Monetary Policy: The Double-Edged Sword of Emergency Response

The interest rate story reads like a tale of good intentions yielding complicated consequences. When COVID-19 struck, the Federal Reserve acted with unprecedented speed and force, dropping rates to essentially zero within weeks. This decisive action prevented a financial system collapse and kept credit flowing during the darkest days of lockdowns. The data shows this emergency response working exactly as intended—markets stabilized, borrowing costs plummeted, and economic activity gradually resumed.

But the same aggressive accommodation that saved the economy in 2020 became its burden by 2022. As inflation surged beyond anyone's expectations, the Fed found itself forced into an equally dramatic reversal. The sharp spike in both short and long-term rates visible in our data represents one of the most aggressive tightening cycles in modern monetary history. What makes this story particularly compelling is how it reveals the inherent tension in central banking: the tools powerful enough to rescue an economy in crisis are often too powerful to use without creating new problems down the line.

```{r}

# Combine data
interest_data <- merge(treasury_10y, fed_funds, by = "DATE", all = TRUE) %>%
  filter(!is.na(Treasury_10Y), !is.na(Fed_Funds_Rate)) %>%
  mutate(Yield_Spread = Treasury_10Y - Fed_Funds_Rate)

# Create simple plotly chart
p <- plot_ly(interest_data, x = ~DATE) %>%
  
  # Add Fed Funds Rate
  add_lines(y = ~Fed_Funds_Rate, name = 'Federal Funds Rate', 
            line = list(color = '#E74C3C', width = 3)) %>%
  
  # Add 10-Year Treasury
  add_lines(y = ~Treasury_10Y, name = '10-Year Treasury', 
            line = list(color = '#3498DB', width = 3)) %>%
  
  # Add yield spread
  add_lines(y = ~Yield_Spread, name = 'Yield Spread', 
            line = list(color = '#95A5A6', width = 2, dash = 'dot')) %>%
    
  # Add COVID-19 vertical line
  add_segments(x = as.Date("2020-03-15"), xend = as.Date("2020-03-15"),
               y = 0, yend = 8, 
               line = list(color = "#2C3E50", width = 2, dash = "dash"),
               name = "COVID-19 Pandemic", showlegend = FALSE) %>%
  
  # Simple layout
  layout(
    title = "Monetary Policy Response: Interest Rates During COVID-19 Era",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Interest Rate (%)"),
    hovermode = 'x'
  )

# Display the plot
p
```

## Energy: The Wild Card in America's Inflation Story

Energy prices have always been the wild card in inflation discussions, but the past few years elevated that volatility to extraordinary levels. The gasoline price data captures two massive shocks that bookend the pandemic era—each telling a different story about how global events land in American driveways. The 2020 collapse in energy prices was so severe that oil futures briefly traded negative, a surreal moment when producers literally paid people to take oil off their hands as storage filled up and demand evaporated.

The subsequent recovery and spike following Russia's invasion of Ukraine demonstrates how quickly energy markets can swing from deflation to inflation. For American families, this volatility translated directly into monthly budget uncertainty. The same commute that cost $40 to fill up in April 2020 suddenly required $70 by March 2022. These aren't abstract numbers—they represent real choices between filling the gas tank and buying groceries, between driving to work and finding remote alternatives, between maintaining normal routines and fundamentally altering daily life.


```{r}
gas_data <- gasoline_cpi %>%
  filter(DATE >= as.Date("2010-01-01"), !is.na(Gasoline_CPI)) %>%
  mutate(
    YoY_Change = ((Gasoline_CPI / lag(Gasoline_CPI, 12)) - 1) * 100,
    Crisis_Period = case_when(
      DATE >= as.Date("2020-01-01") & DATE <= as.Date("2021-12-31") ~ "COVID-19 Pandemic",
      DATE >= as.Date("2022-01-01") ~ "Post-COVID Inflation",
      TRUE ~ "Normal Period"
    )
  )

# Create subplot with multiple perspectives
energy_plotly <- subplot(
  # Main gasoline CPI plot
  plot_ly(gas_data, x = ~DATE, y = ~Gasoline_CPI, type = 'scatter', mode = 'lines',
          line = list(color = '#E67E22', width = 3), name = 'Gasoline CPI') %>%
    add_markers(data = subset(gas_data, Crisis_Period != "Normal Period"),
                x = ~DATE, y = ~Gasoline_CPI, color = ~Crisis_Period,
                colors = c('#3498DB', '#F39C12'),
                marker = list(size = 4, opacity = 0.7),
                name = 'Crisis Periods') %>%
    layout(yaxis = list(title = "Gasoline CPI")),
  
  # Year-over-year change plot
  plot_ly(gas_data, x = ~DATE, y = ~YoY_Change, type = 'scatter', mode = 'lines',
          line = list(color = '#8E44AD', width = 2), name = 'YoY % Change') %>%
    add_trace(x = range(gas_data$DATE, na.rm = TRUE), y = c(0, 0), 
              type = 'scatter', mode = 'lines',
              line = list(dash = "dash", color = "gray", width = 1),
              name = 'Zero Line', showlegend = FALSE) %>%
    layout(yaxis = list(title = "YoY Change (%)")),
  
  nrows = 2, heights = c(0.7, 0.3), shareX = TRUE
) %>%
  
  layout(
    title = list(text = "<b>Energy Price Volatility During Economic Disruptions</b><br><sub>How COVID-19 and Geopolitical Events Shaped Gasoline Costs</sub>",
                 font = list(size = 16)),
    xaxis = list(title = "Year", 
                 tickmode = "linear",
                 dtick = "M24",
                 tickformat = "%Y"),
    showlegend = TRUE,
    annotations = list(
      list(x = as.Date("2020-04-01"), y = 150, text = "Demand Collapse<br>Negative Oil Prices",
           showarrow = FALSE, font = list(size = 12, color = "#2C3E50"),
           bgcolor = "rgba(255,255,255,0.8)", bordercolor = "#3498DB", borderwidth = 1,
           xref = "x", yref = "y"),
      list(x = as.Date("2022-03-01"), y = 380, text = "Russia-Ukraine<br>Supply Disruption",
           showarrow = FALSE, font = list(size = 12, color = "#2C3E50"),
           bgcolor = "rgba(255,255,255,0.8)", bordercolor = "#F39C12", borderwidth = 1,
           xref = "x", yref = "y")
    )
  )

# Display the energy plot
energy_plotly

```

What emerges from the above three visualizations is a portrait of an economy in transition, where traditional patterns have been disrupted and new realities are still taking shape. The COVID-19 pandemic didn't simply pause normal economic life—it fundamentally altered the relationship between monetary policy, asset prices, and household budgets. Housing became less affordable even as borrowing became cheaper, energy prices swung between extremes with little warning, and families found themselves navigating an economic landscape where past experience provided little guidance for future planning.

These changes represent more than statistical variations; they reflect altered life trajectories, deferred dreams, and adapted expectations. The young couple saving for their first home discovers that their target keeps moving further away. The family planning a budget finds that energy costs make mockery of careful calculations. The retiree on fixed income watches purchasing power erode as the very policies meant to support the economy inflate away their savings.

The data tells us that we're living through a period where economic policy has become more powerful and more consequential than ever before. The tools that central banks and governments wield can prevent depressions, but they can also accidentally create asset bubbles, affordability crises, and wealth inequality. Understanding these patterns isn't just an academic exercise—it's essential for anyone trying to navigate an economy where the old rules no longer seem to apply.
