---
title: "Univariate Time Series Models"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 4
    link-external-newwindow: true
code-fold: true
code-summary: "Show Code"
execute:
  warning: false
  message: false
editor : source
---

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages({
  library(readr); library(dplyr); library(zoo); library(lubridate)
  library(forecast); library(tseries); library(ggplot2); library(knitr);library(astsa)
})

fred_read <- function(path){
  read_csv(path, show_col_types = FALSE) |>
    rename(date = 1, value = 2) |>
    mutate(date = as.Date(date), value = suppressWarnings(as.numeric(value))) |>
    filter(!is.na(value))
}

to_monthly <- function(df, fun = mean){
  df |>
    mutate(ym = as.yearmon(date)) |>
    group_by(ym) |>
    summarize(value = fun(value, na.rm = TRUE), .groups = "drop") |>
    arrange(ym)
}

ym_start <- function(df){
  y <- as.numeric(min(df$ym)); y_int <- floor(y); m <- round((y - y_int) * 12) + 1
  c(y_int, m)
}

safe_arima <- function(x, order, seasonal = NULL, include.drift = FALSE) {
  # Try ML; if it fails, fall back to CSS-ML; if still fails, return NULL
  tryCatch({
    Arima(x, order = order, seasonal = seasonal,
          include.drift = include.drift, method = "ML")
  }, error = function(e) {
    tryCatch({
      Arima(x, order = order, seasonal = seasonal,
            include.drift = include.drift, method = "CSS-ML")
    }, error = function(e2) {
      NULL
    })
  })
}

grid_arima <- function(x, d = 1, pmax = 4, qmax = 4, seasonal = NULL) {
  # Robust grid search that skips unstable models and never stops the render
  n <- (pmax + 1) * (qmax + 1)
  rows <- vector("list", n)
  k <- 1
  for (p in 0:pmax) {
    for (q in 0:qmax) {
      fit <- safe_arima(
        x,
        order = c(p, d, q),
        seasonal = seasonal,
        include.drift = (d == 1)
      )
      if (is.null(fit)) {
        rows[[k]] <- data.frame(
          p = p, d = d, q = q,
          AIC = NA_real_, BIC = NA_real_, AICc = NA_real_,
          status = "failed"
        )
      } else {
        rows[[k]] <- data.frame(
          p = p, d = d, q = q,
          AIC = AIC(fit),
          BIC = BIC(fit),
          AICc = if (!is.null(fit$aicc)) fit$aicc else NA_real_,
          status = "ok"
        )
      }
      k <- k + 1
    }
  }
  out <- dplyr::bind_rows(rows)
  out[order(out$AIC), ]
}
```

# Data Loading & Preprocessing

```{r}
# Paths
path <- "data"
# CPI (headline + food)
fh   <- to_monthly(fred_read(file.path(path, "CUSR0000SAF11.csv"))) # Food at Home
fa   <- to_monthly(fred_read(file.path(path, "CUSR0000SEFV.csv")))  # Food Away from Home
cpi  <- to_monthly(fred_read(file.path(path, "CPIAUCSL.csv")))      # Overall CPI
# Rates
dgs  <- to_monthly(fred_read(file.path(path, "DGS10.csv")))         # 10Y Treasury
ff   <- to_monthly(fred_read(file.path(path, "FEDFUNDS.csv")))      # Fed Funds

# ts objects (monthly)
fh_ts  <- ts(fh$value,  start = ym_start(fh),  frequency = 12)
fa_ts  <- ts(fa$value,  start = ym_start(fa),  frequency = 12)
cpi_ts <- ts(cpi$value, start = ym_start(cpi), frequency = 12)
dgs_ts <- ts(dgs$value, start = ym_start(dgs), frequency = 12)
ff_ts  <- ts(ff$value,  start = ym_start(ff),  frequency = 12)

# log-transform for CPI-like series
fh_log  <- log(fh_ts)
fa_log  <- log(fa_ts)
cpi_log <- log(cpi_ts)
```

> **Note.** CPI series are modeled on the **log scale** (common in inflation work). Interest rates remain on levels.

---

## Overall CPI (Headline) 

::: {.panel-tabset}

### ACF & PACF {.active}

```{r}
# ACF/PACF on differenced log CPI (d=1, seasonal D=1 at s=12)
cpid <- diff(diff(cpi_log, 1), 12)
set.seed(123)
GG1 <- ggAcf(cpid) + ggtitle("ACF: Δ & seasonal Δ log(CPI)")
GG2 <- ggPacf(cpid) + ggtitle("PACF: Δ & seasonal Δ log(CPI)")
GG1; GG2
```

**Takeaway.** Clear seasonal pattern and lag-1 behavior → try **SARIMA(0,1,1)(0,1,1)[12]**.

### Manual ARIMA

```{r}
# Grid over p,q with seasonal part fixed at (0,1,1)[12]
manual_cpi <- grid_arima(cpi_log, d = 1, pmax = 4, qmax = 4,
                         seasonal = list(order = c(0,1,1), period = 12))
knitr::kable(head(manual_cpi, 10), digits = 3)
```

**Reading the table.** Lower AIC/BIC indicates better fit; top rows are the leading candidates.

### Equation

The airline form (backshift \(L\)):

$$
(1-L)(1-L^{12})\,y_t \;=\; (1+0.5817\,L)\,(1-0.9211\,L^{12})\,\varepsilon_t,
\qquad y_t=\log(\mathrm{CPI}).
$$

Expanded right-hand side:

$$
(1-L)(1-L^{12})\,y_t
=\big(1 + 0.5817\,L - 0.9211\,L^{12} - 0.5358\,L^{13}\big)\,\varepsilon_t,
$$

where $0.5358 \approx 0.9211\times 0.5817.$

```{r}
astsa::sarima(cpi_log, 0, 1, 1,  0, 1, 1,  12)
```

### auto.arima()

```{r}
auto_cpi <- auto.arima(cpi_log, seasonal = TRUE)
auto_cpi
```

**Comment.** auto.arima may propose close variants (e.g., (0,2,2)); prefer the model with the best diagnostics and lowest IC.

### Forecast

```{r}
fit_cpi <- Arima(cpi_log, order = c(0,1,1), seasonal = list(order = c(0,1,1), period = 12))
fc_cpi  <- forecast(fit_cpi, h = 12)
autoplot(fc_cpi) + ggtitle("Forecast: Overall CPI (log)")
```

**Interpretation.** Short-run uncertainty is modest; bands widen at longer horizons.

### Benchmark Comparison

```{r}
autoplot(cpi_log) +
  autolayer(meanf(cpi_log, h = 12),  series = "Mean",     PI = FALSE) +
  autolayer(naive(cpi_log, h = 12),  series = "Naive",    PI = FALSE) +
  autolayer(snaive(cpi_log, h = 12), series = "Seasonal", PI = FALSE) +
  autolayer(fc_cpi, series = "SARIMA", PI = FALSE) +
  ggtitle("Benchmarks vs SARIMA: log(CPI)") + xlab("Time") + ylab("log(CPI)")
```

**Result.** SARIMA should outperform Seasonal Naive in RMSE/MAE; if not, revisit seasonal terms.

::: 

---

## Food at Home CPI 

::: {.panel-tabset}

### ACF & PACF {.active}

```{r}
fh_d <- diff(diff(fh_log, 1), 12)
set.seed(123)
GG1 <- ggAcf(fh_d) + ggtitle("ACF: Δ & seasonal Δ log(Food at Home)")
GG2 <- ggPacf(fh_d) + ggtitle("PACF: Δ & seasonal Δ log(Food at Home)")
GG1; GG2
```

**Takeaway.** Seasonality is strong; an airline-style model **SARIMA(0,1,1)(0,1,1)[12]** is a natural baseline.

### Manual ARIMA

```{r}
manual_fh <- grid_arima(fh_log, d = 1, pmax = 4, qmax = 4,
                        seasonal = list(order = c(0,1,1), period = 12))
knitr::kable(head(manual_fh, 10), digits = 3)
```

### Equation

The airline form (backshift \(L\)):

$$
(1-L)(1-L^{12})\,y_t \;=\; (1+0.3428\,L)\,(1-0.9994\,L^{12})\,\varepsilon_t,
\qquad y_t=\log(\mathrm{Food\ at\ Home\ CPI}).
$$

Expanded right-hand side:

$$
(1-L)(1-L^{12})\,y_t
=\big(1 + 0.3428\,L - 0.9994\,L^{12} - 0.3426\,L^{13}\big)\,\varepsilon_t,
$$

where $0.3426 \approx 0.9994\times 0.3428.$

```{r}
# Food at Home (log), SARIMA(0,1,1)(0,1,1)[12]
astsa::sarima(fh_log, 0, 1, 1,  0, 1, 1,  12)

```

### auto.arima()

```{r}
auto_fh <- auto.arima(fh_log, seasonal = TRUE)
auto_fh
```

### Forecast

```{r}
fit_fh <- Arima(fh_log, order = c(0,1,1), seasonal = list(order = c(0,1,1), period = 12))
fc_fh  <- forecast(fit_fh, h = 12)
autoplot(fc_fh) + ggtitle("Forecast: Food at Home CPI (log)")
```

### Benchmark Comparison

```{r}
autoplot(fh_log) +
  autolayer(snaive(fh_log, h = 12), series = "Seasonal", PI = FALSE) +
  autolayer(fc_fh, series = "SARIMA", PI = FALSE) +
  ggtitle("Benchmarks vs SARIMA: log(Food at Home)")
```

::: 

---

## Food Away from Home CPI 

::: {.panel-tabset}

### ACF & PACF {.active}

```{r}
fa_d <- diff(diff(fa_log, 1), 12)
set.seed(123)
GG1 <- ggAcf(fa_d) + ggtitle("ACF: Δ & seasonal Δ log(Food Away)")
GG2 <- ggPacf(fa_d) + ggtitle("PACF: Δ & seasonal Δ log(Food Away)")
GG1; GG2
```

### Manual ARIMA

```{r}
manual_fa <- grid_arima(fa_log, d = 1, pmax = 4, qmax = 4,
                        seasonal = list(order = c(0,1,1), period = 12))
knitr::kable(head(manual_fa, 10), digits = 3)
```

### Equation

The airline form (backshift \(L\)):

$$
(1-L)(1-L^{12})\,y_t \;=\; (1+0.5485\,L)\,(1-0.6023\,L^{12})\,\varepsilon_t,
\qquad y_t=\log(\mathrm{Food\ Away\ from\ Home\ CPI}).
$$

Expanded right-hand side:

$$
(1-L)(1-L^{12})\,y_t
=\big(1 + 0.5485\,L - 0.6023\,L^{12} - 0.3304\,L^{13}\big)\,\varepsilon_t,
$$

where $0.3304 \approx 0.5485\times 0.6023.$

```{r}
# Food Away (log), SARIMA(0,1,1)(0,1,1)[12]
astsa::sarima(fa_log, 0, 1, 1,  0, 1, 1,  12)

```

### auto.arima()

```{r}
auto_fa <- auto.arima(fa_log, seasonal = TRUE)
auto_fa
```

### Forecast

```{r}
fit_fa <- Arima(fa_log, order = c(0,1,1), seasonal = list(order = c(0,1,1), period = 12))
fc_fa  <- forecast(fit_fa, h = 12)
autoplot(fc_fa) + ggtitle("Forecast: Food Away CPI (log)")
```

### Benchmark Comparison

```{r}
autoplot(fa_log) +
  autolayer(snaive(fa_log, h = 12), series = "Seasonal", PI = FALSE) +
  autolayer(fc_fa, series = "SARIMA", PI = FALSE) +
  ggtitle("Benchmarks vs SARIMA: log(Food Away)")
```

::: 

---

## 10-Year Treasury Yield (DGS10) 

::: {.panel-tabset}

### ACF & PACF {.active}

```{r}
dgs_d <- diff(dgs_ts)
set.seed(123)
GG1 <- ggAcf(dgs_d) + ggtitle("ACF: Δ 10Y Treasury")
GG2 <- ggPacf(dgs_d) + ggtitle("PACF: Δ 10Y Treasury")
GG1; GG2
```

**Takeaway.** Short memory with lag-1 prominence → **ARIMA(0,1,1)** works well.

### Manual ARMA


```{r}

suppressPackageStartupMessages(library(forecast))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(dplyr))

as_monthly_ts <- function(x) {

  if (inherits(x, "ts")) {
    x <- stats::na.omit(x)
    ts(as.numeric(x), start = start(x), frequency = ifelse(frequency(x) > 1, frequency(x), 12))
  } else {
    x <- stats::na.omit(as.numeric(x))
    ts(x, frequency = 12)
  }
}

dgs_ts <- as_monthly_ts(dgs_ts)
ff_ts  <- as_monthly_ts(ff_ts)

stopifnot(frequency(dgs_ts) == 12, frequency(ff_ts) == 12)

safe_arima_fit <- function(x, p, d, q, include_drift = (d == 1)) {
  fit <- tryCatch(
    Arima(x, order = c(p, d, q), include.drift = include_drift, method = "ML"),
    error = function(e) NULL
  )
  if (is.null(fit)) {
    fit <- tryCatch(
      Arima(x, order = c(p, d, q), include.drift = include_drift, method = "CSS-ML"),
      error = function(e) NULL
    )
  }
  fit
}

manual_arima_grid <- function(x, d = 1, pmax = 4, qmax = 4) {
  rows <- vector("list", (pmax + 1) * (qmax + 1))
  errs <- vector("list", length(rows))
  k <- 1
  for (p in 0:pmax) {
    for (q in 0:qmax) {
      fit <- safe_arima_fit(x, p, d, q)
      if (is.null(fit)) {
        rows[[k]] <- data.frame(p = p, d = d, q = q,
                                AIC = NA_real_, BIC = NA_real_, AICc = NA_real_,
                                status = "failed")
      } else {
        rows[[k]] <- data.frame(p = p, d = d, q = q,
                                AIC = AIC(fit),
                                BIC = BIC(fit),
                                AICc = if (!is.null(fit$aicc)) fit$aicc else NA_real_,
                                status = "ok")
      }
      k <- k + 1
    }
  }
  out <- do.call(rbind, rows)
  out <- out[order(out$BIC), ]
  rownames(out) <- NULL
  out
}

grid_dgs <- manual_arima_grid(dgs_ts, d = 1, pmax = 4, qmax = 4)

kable(head(grid_dgs, 10), digits = 3, caption = "DGS10 — Manual ARIMA(p,1,q) (sorted by BIC)")

```

### Equation

Backshift form:

$$
(1-L)\,y_t \;=\; 0.0036 \;+\; (1+0.3690\,L)\,\varepsilon_t,
\qquad y_t=\text{10Y Treasury}.
$$

Expanded right-hand side:

$$
(1-L)\,y_t \;=\; 0.0036 \;+\; \big(1+0.3690\,L\big)\,\varepsilon_t.
$$

*(Here the constant 0.0036 is the drift in the differenced series.)*

```{r}
# 10Y Treasury, ARIMA(0,1,1)
astsa::sarima(dgs_ts, 0, 1, 1)

```

### auto.arima()

```{r}
auto_dgs <- auto.arima(dgs_ts)
auto_dgs
```

### Forecast

```{r}
fit_dgs <- Arima(dgs_ts, order = c(0,1,1))
fc_dgs  <- forecast(fit_dgs, h = 12)
autoplot(fc_dgs) + ggtitle("Forecast: 10Y Treasury")
```

### Benchmark Comparison

```{r}
autoplot(dgs_ts) +
  autolayer(meanf(dgs_ts, h = 12), series = "Mean", PI = FALSE) +
  autolayer(naive(dgs_ts, h = 12), series = "Naive", PI = FALSE) +
  autolayer(fc_dgs, series = "ARIMA", PI = FALSE) +
  ggtitle("Benchmarks vs ARIMA: 10Y Treasury")
```

::: 

---

## Federal Funds Rate 

::: {.panel-tabset}

### ACF & PACF {.active}

```{r}
ff_d <- diff(ff_ts)
set.seed(123)
GG1 <- ggAcf(ff_d) + ggtitle("ACF: Δ Fed Funds")
GG2 <- ggPacf(ff_d) + ggtitle("PACF: Δ Fed Funds")
GG1; GG2
```

**Takeaway.** Persistence at lag-1 → **ARIMA(1,1,0)** is a solid baseline.

### Manual ARMA

```{r}

grid_ff  <- manual_arima_grid(ff_ts,  d = 1, pmax = 4, qmax = 4)
kable(head(grid_ff,  10), digits = 3, caption = "Fed Funds — Manual ARIMA(p,1,q) (sorted by BIC)")

```

### Equation

Backshift form:

$$
(1-0.6574\,L)\,(1-L)\,y_t \;=\; 0.0224 \;+\; \varepsilon_t,
\qquad y_t=\text{Fed Funds}.
$$

Expanded left-hand operator:

$$
\big(1 - L - 0.6574\,L + 0.6574\,L^{2}\big)\,y_t \;=\; 0.0224 \;+\; \varepsilon_t.
$$

```{r}
# Federal Funds, ARIMA(1,1,3)
astsa::sarima(ff_ts, 1, 1, 3)
```

### auto.arima()

```{r}
auto_ff <- auto.arima(ff_ts)
auto_ff
```

### Forecast

```{r}
fit_ff <- Arima(ff_ts, order = c(1,1,0))
fc_ff  <- forecast(fit_ff, h = 12)
autoplot(fc_ff) + ggtitle("Forecast: Federal Funds Rate")
```

### Benchmark Comparison

```{r}
autoplot(ff_ts) +
  autolayer(meanf(ff_ts, h = 12), series = "Mean", PI = FALSE) +
  autolayer(naive(ff_ts, h = 12), series = "Naive", PI = FALSE) +
  autolayer(fc_ff, series = "ARIMA", PI = FALSE) +
  ggtitle("Benchmarks vs ARIMA: Fed Funds")
```

::: 